---
title: "Abundance analyses"
author: "Jeff Oliver"
date: "August 23, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Testing for effects of sampling method on abundances, as well as how size influences abundance

# Methods
## Setup
Load data and third-party packages.

```{r data-setup}
library("tidyr")   # for gather
library("ggplot2") # for plotting
library("nlme")    # for mixed-effect models (lme)
bioscan <- read.csv(file = "data/BioScanData.csv")
species.data <- read.csv(file = "data/species-data.csv")

# Drop any rows missing data
bioscan <- na.omit(bioscan)
```

## Data wrangling
Only want to use data for those sites where we have information for both collection types.

```{r identify-complete-data}
# Identify sites with data for each of the two collection methods
pollard.sites <- bioscan$Site.Number[bioscan$Collection.Method == "Pollard Walk"]
malaise.sites <- bioscan$Site.Number[bioscan$Collection.Method == "Malaise"]

# Identify sites with data for *both* collection methods
sites.with.both <- intersect(x = pollard.sites, y = malaise.sites)

# Reduce dataset to only those sites with both types of data
bioscan <- bioscan[bioscan$Site.Number %in% sites.with.both, ]
rownames(bioscan) <- NULL
```

We also need to transform our data so (1) each row corresponds to a single species and (2) the species size data is also included.

```{r data-transformation}
# Transform data to long format
bioscan.long <- bioscan %>%
  gather(key = "Species", value = "Abundance", -c(1:4))

# Add species data
bioscan.long <- merge(x = bioscan.long, 
                      y = species.data, 
                      by.x = "Species", 
                      by.y = "species")
# Use the mean of the minimum and maximum as single value for size
bioscan.long$size <- apply(X = bioscan.long[, c("opler.wright.min", "opler.wright.max")],
                           MARGIN = 1,
                           FUN = function(x) {mean(x)})
```

## Statistical analysis

### Effect of sampling type
Test the effect of collection method on abundance, including site as a random intercept effect.

$$
Abundance = \beta_0 + \beta_1 \times Collection Method + b_{0i}
$$

where $b_{0i}$ is a random intercept for the _i^th^_ site.


```{r test-non-zero-intercept}
# Run simple model, testing the null hypothesis of intercept = 0 (no difference
# in abundance)
simple.model <- lme(Abundance ~ Collection.Method, 
                    random = ~1|Site.Number, 
                    data = bioscan.long)
simple.tTable <- summary(simple.model)$tTable
```

### Effect of species size
Test a model where size, as well as the interaction between collection method and size, are included in the model to predict abundance.

$$
Abundance = \beta_0 + \beta_1 \times Collection Method + \beta_2 \times Size + \beta_3 \times Collection Method \times Size + b_{0i}
$$

where $b_{0i}$ is a random intercept for the _i^th^_ site.

```{r test-size-effect}
# Run a more complex model, with size as a fixed effect, as well as the 
# interaction between size and Collection.Method
size.model <- lme(Abundance ~ Collection.Method + size + Collection.Method*size, 
                  random = ~1|Site.Number, 
                  data = bioscan.long)
size.tTable <- summary(size.model)$tTable
```

## Results
```{r p-value-text, echo = FALSE}
# Have to do some formatting if p-value is really low, otherwise it just gets 
# reported as "0"
simple.p.value <- "< 0.001"
if (simple.tTable[2, "p-value"] > 0.001) {
  simple.p.value <- paste0("= ", round(simple.tTable[2, "p-value"], 3))
}
```

On average, abundance was higher in Pollard walk surveys than in Malaise traps (_t_ = `r round(simple.tTable[2, "t-value"], 3)`, p `r simple.p.value`). I'm not sure this plot is the best way to visualize this...
```{r abundance-boxplot, out.height = "250px"}
abundance.distribution <- ggplot(data = bioscan.long, 
                                 mapping = aes(x = Abundance, fill = Collection.Method)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  xlab(label = "Abundance") +
  ylab(label = "Frequency") +
  scale_fill_manual(values = c("#CA3542", "#27647B"), name = "Collection Method") +
  theme_bw()
print(abundance.distribution)
```

Neither species size nor the interaction between size and collection method have a significant effect on abundance:

```{r round-size-coefficients, echo = FALSE}
size.tTable <- round(size.tTable, 3)
```

| Coefficient | Estimate | _t_ | _p_ |
|:------------|:---:|:---:|:---:|
| Intercept   | `r size.tTable[1, "Value"]` | `r size.tTable[1, "t-value"]` | `r size.tTable[1, "p-value"]` |
| Collection method | `r size.tTable[2, "Value"]` | `r size.tTable[2, "t-value"]` | `r size.tTable[2, "p-value"]` |
| Size | `r size.tTable[3, "Value"]` | `r size.tTable[3, "t-value"]` | `r size.tTable[3, "p-value"]` |
| Collection method x Size | `r size.tTable[4, "Value"]` | `r size.tTable[4, "t-value"]` | `r size.tTable[4, "p-value"]` |

