---
title: "Frequency Distributions"
author: "Jeff Oliver"
date: "September 20, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare the frequency of species' occurence across sites in the two collection methods

```{r load-dependencies}
library("ggplot2") # plotting 
library("dplyr")
library("tidyr")   # for long-data formatting
source(file = "bioscan-functions.R")
bioscan <- CompleteBioscan()
```

Want to create a histogram with:

+ X-axis: the number of sites
+ Y-axis: number of species

Shows the frequency distribution of species, to see if there were lots of singlets/doublets, or if there were many species seen at many sites.

We'll want a data frame like:

| Collection Method | Species              | # Sites |
|:------------------|:---------------------|:-------:|
| Malaise           | _Agraulis\_vanillae_ | 6       |
| Malaise           | _Brephidium\_exilis_ | 1       |
| ...               | ...                  | ...     |
| Pollard Walk      | _Vanessa\_cardui_    | 13      |


```{r calculate-frequencies}
# Identify those columns with species data
species.cols <- c(5:33)

# Count just whether or not a species occurred for a particular site/method
bioscan.freqs <- as.data.frame(bioscan[, species.cols] > 0)
bioscan.freqs$Collection.Method <- bioscan$Collection.Method

# There has to be a tidyverse way of doing this, but for now, brute force
# Separate the two collection methods
malaise.freqs <- bioscan.freqs[bioscan.freqs$Collection.Method == "Malaise", ]
pollard.freqs <- bioscan.freqs[bioscan.freqs$Collection.Method == "Pollard Walk", ]

# Use boolean math to count the number of sites each species occurred in
malaise.sums <- colSums(malaise.freqs[, -(which(colnames(malaise.freqs) == "Collection.Method"))])
pollard.sums <- colSums(pollard.freqs[, -(which(colnames(pollard.freqs) == "Collection.Method"))])

# Convert to long format and add method as new column
malaise.long <- data.frame(species = names(malaise.sums),
                           num.sites = malaise.sums)
malaise.long$Collection.Method <- "Malaise"
pollard.long <- data.frame(species = names(pollard.sums),
                           num.sites = pollard.sums)
pollard.long$Collection.Method <- "Pollard Walk"

# Combine data back together for plot
bioscan.long <- rbind(malaise.long, pollard.long)
# Drop any zeros, since we're not interested in plotting those
bioscan.long <- bioscan.long[bioscan.long$num.sites > 0, ]
rownames(bioscan.long) <- NULL
```

We now have a table, so we can plot the histogram.

```{r plot-freqs}
freq.plot <- ggplot(data = bioscan.long, mapping = aes(x = num.sites)) +
  geom_histogram(bins = max(bioscan.long$num.sites)) +
  facet_wrap(~ Collection.Method) +
  theme_bw() +
  xlab(label = "Number of Sites") +
  ylab(label = "Number of Species") +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16))
print(freq.plot)
ggsave(filename = "output/frequency-distribution.png", plot = freq.plot)
```