---
title: "Frequency Distributions"
author: "Jeff Oliver"
date: "September 20, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Compare the frequency of species' occurence across sites in the two collection methods

```{r load-dependencies}
library("ggplot2") # plotting 
library("dplyr")
library("tidyr")   # for long-data formatting
source(file = "bioscan-functions.R")
bioscan <- CompleteBioscan()
```

Want to create a histogram with:

+ X-axis: the number of sites
+ Y-axis: number of species

Shows the frequency distribution of species, to see if there were lots of singlets/doublets, or if there were many species seen at many sites.

We'll want a data frame like:

| Collection Method | Species | # Sites |
|:------------------|:--------|:-------:|
| Malaise           | _Agraulis\_vanillae_ |  |


```{r calculate-frequencies}
# Identify those columns with species data
species.cols <- c(5:33)

# Count just whether or not a species occurred for a particular site/method
bioscan.freqs <- as.data.frame(bioscan[, species.cols] > 0)
bioscan.freqs$Collection.Method <- bioscan$Collection.Method

bioscan.sums <- bioscan.freqs %>%
  group_by(Collection.Method)
```
